Effective Dynamic Models of Metabolic Networks

### 2016-03-13 Paper_v1.tex
Results section done

The performance of HCM FBA was equivalent to HCM EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
The anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, generated nine elementary and seven FBA modes.
HCM EM cellmass, glucose and fermentation product trajectories were generated
using these elementary modes and kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, dashed).
We estimated the HCM FBA model parameters from the experimental data set taken from Kim et al.  \cite{2008_kim_varner_ramkrishna_BiotechProg} using simulated annealing.
Overall, HCM FBA performed only 5\% worse than HCM EM (Fig. \ref {fig:ecoli}A, solid), despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
Taken together, these results suggested HCM FBA is an alternative to HCM EM even for cases when elementary modes can be easily calculated.

@@

HCM FBA outperformed HCM EM for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
An \emph{E. coli} metabolic network, consisting of 60 metabolites and 110 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
Elementary mode decomposition of this network was not feasible; we generated $>$ 150,000 elementary modes before prematurely terminating the calculation.
Conversely, flux balance analysis generated only 29 modes for the same network.
HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements generated by Palsson and coworkers \cite{1994_varma_palsson_ApplEnvMicro} (Fig. \ref{fig:ecoli}B, solid).
HCM FBA captured glucose consumption, cellmass formation and the switch between acetate production and consumption following glucose exhaustion.
Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA was able to describe cell growth, including the switch between glucose and acetate consumption,
for a network that was infeasible to decompose using elementary modes.
However, it was unclear whether all 29 FBA modes were required to capture the experimental data.

Variance based sensitivity of aerobic \textit{E. coli} metabolism identified which FBA modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions.
We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.
Five FBA modes were determined to be significant by an arbitrary threshold of 0.00218.
The remaining 24 FBA modes were identified to be minor modes and insignificant.
The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B, dotted)
was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.

>>>>

The performance of HCM FBA was equivalent to HCM EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated nine elementary and seven FBA modes.
HCM EM cellmass, glucose and fermentation product trajectories, generated
using kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg},
were compared with experimental measurements (Fig. \ref{fig:ecoli}A, dashed).
We estimated HCM FBA model parameters from the Kim et al. experimental data set using simulated annealing.
Overall, HCM FBA performed within 5\% of HCM EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
Taken together, these results suggested HCM FBA  was capable of describing experimental data, and is a viable reduced-order alternative to HCM EM.

@@

HCM FBA captured the glucose to acetate metabolic shift for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
A core \emph{E. coli} metabolic network, consisting of 60 metabolites and 105 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
Elementary mode decomposition (and thus HCM EM) was not feasible for this network; 153,000 elementary modes were generated before the calculation prematurely terminated.
Conversely, flux balance analysis generated only 29 modes for the same network.
HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
HCM FBA captured glucose consumption, cellmass formation and the dynamic switch between acetate production and consumption following glucose exhaustion.
Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA described metabolic dynamics for large
networks that were infeasible for HCM EM. However, it was unclear whether all 29 FBA modes were required to capture the experimental data.
Next, we considered a systematic approach to identify the critical subset of FBA modes required for model performance.

Global variance based sensitivity analysis of aerobic \textit{E. coli} metabolism identified the FBA modes essential to biomass yield (Fig. \ref{fig:sensitivity}).
Total order sensitivity coefficients were calculated for all kinetic parameters and enzyme initial conditions.
Five of the 29 FBA modes were determined to be significant by a threshold test.
Removal of the most significant mode (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
Conversely, removal of the remaining 24 modes had negligible effect on model performance (Fig. \ref{fig:ecoli}B, dashed).
Taken together, these results demonstrate that sensitivity analysis can identify critical FBA modes, thereby allowing further reduction to the minimal model structure
required to explain the data.
